A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
rework on temporal processing

-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time decomposed into several components, such as year, month, quarter, and season.

@@

| No. | Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
|-----|-----------------|----------|--------------------|---------------------|
| 1   | Temporal processing     | $x(s;t)$          | aaaaa                                                                                                                                              | $x(s; t^\prime)$                                                    |
| 2   | Spatial processing      | $x(s;t)$          | aaaaa                                                                                                                                              |  $x(s^\prime; t)$                                                   |
| 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????                                                       |
| 4   | Scaling                 | $x(s; t)$         | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
| 6   | Distribution fit        | $x(s; t)$         | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1 |
| 7   | Normalising             | $x(s; t)$         | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                     |
| 8   | Benchmarking            | $x(s; t)$         |                                                                                                                                                    | $b(s;t)$                                                            |
| 9   | Simplification          | $x(s; t)$         | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                              |

@@

`Output`: $x(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

`Computation`: The construction of an index sometimes needs to consider information from neighbouring time periods. The temporal processing is a general operator on the time dimension of the data in the form of

```{=tex}
\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}
```
where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. A typical example of temporal processing is aggregation and in the drought community, these indexes are often called multi-scalar indexes. For example, Standardized Precipitation Index (SPI) aggregates monthly precipitation using a time scale parameter $k$ in the form of

$$x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),$$

where $j^\prime$ is the new time index after the aggregation. Multi-scalar drought indexes are widely used to study different types of drought events. A smaller value of the time scales parameter $k$ is used for short term drought events, while a large $k$ of 18 or 24, an equivalent to a 1.5- or 2-year aggregation, is often used for long term assessment.

>>>>

-   $t$ denotes the temporal order in $D_t \subseteq \mathbb{R}^n$. Sometimes, it can be time breaks into other temporal components, such as year, month, quarter, and season.

@@

| No. | Module                  | Input             | Operation                                                                                                                                          | Output                                                              |
|-------------|-------------|-------------|-----------------|-----------------|
| 1   | Temporal processing     | $x(t)$            | $f[x(.)]$                                                                                                                                          | $x^*(t^\prime)$ where $t\prime \in D_{t^\prime}$                    |
| 2   | Spatial processing      | $x(s)$            | aaaaa                                                                                                                                              | $x^\#(s^\prime)$                                                    |
| 3   | Variable transformation | $x(s; t)$         | $T(x(s;t))$                                                                                                                                        | $u(s;t)$ ????                                                       |
| 4   | Scaling                 | $x(s; t)$         | $[x(s;t) - \alpha]/\gamma$                                                                                                                         | $v(s;t)$ where bbbbb                                                |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$ | aaaaa                                                                                                                                              | $\mathbf{y}(s;t)$ where $y \in \mathbb{R}^d$ and $d < p$            |
| 6   | Distribution fit        | $x(s; t)$         | $F_\mathbf{\theta}(x(s,t))$ where $F_{\mathbf{\theta}}(.)$ is the cumulated distribution function with distribution parameter(s) $\mathbf{\theta}$ | $p(s,t)$ where $p(.)$ denotes the probability value between 0 and 1 |
| 7   | Normalising             | $x(s; t)$         | $\Phi^{-1}[x(s, t)]$                                                                                                                               | $z(s; t)$ where $z \sim N(.,.)$                                     |
| 8   | Benchmarking            | $x(s; t)$         |                                                                                                                                                    | $b(s;t)$                                                            |
| 9   | Simplification          | $x(s; t)$         | ccccc                                                                                                                                              | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                              |

@@

`Output`: $x^*(t^{\prime})$, where $t^\prime \in D_{t^{\prime}}$.

`Computation`: The temporal processing module takes a single variable in a set of fixed locations, $x(t)$, as the input. This step *transforms* the original series into a new series to incorporate the temporal neighbouring effect: $x^*(.) = f[x(.)]$. In temporal processing, the temporal dimension $t$ may also transition to $t^\prime \subseteq D_{t^\prime}$, allowing for the decomposition of time into its constituent temporal components. For example, this can be used to break down daily data into monthly. This gives the output of the temporal processing module as $x^*(t^\prime)$.

An example of temporal processing is in Standardized Precipitation Index (SPI), where monthly precipitation is summed over a rolling window of size $k$, also known as the time scale. Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be written as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$. For SPI, the choice of the time scale $k$, is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

<!-- The construction of an index sometimes needs to consider information from neighbouring time periods. The temporal processing is a general operator on the time dimension of the data in the form of -->

<!-- ```{=tex} -->

<!-- x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})), -->

<!-- ``` -->

<!-- where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. -->

### 2023-08-15 tidyindex.tex
rework on temporal processing

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^d, t \in D_t \subseteq \mathbb{R}^m 
\end{equation}

@@

\begin{itemize}
\item
  \(x\) is the variable in \(\mathbb{R}^p\),
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^d\). This includes a collection of
  countries of interest and paired longitude and latitude coordinates of
  locations, and
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^m\).
  Sometimes, it can be time decomposed into several components, such as
  year, month, quarter, and season.
\end{itemize}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0685}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1370}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2740}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2877}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(s;t)\) & aaaaa & \(x(s; t^\prime)\) \\
2 & Spatial processing & \(x(s;t)\) & aaaaa & \(x(s^\prime; t)\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Output}: \(x(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

\texttt{Computation}: The construction of an index sometimes needs to
consider information from neighbouring time periods. The temporal
processing is a general operator on the time dimension of the data in
the form of

\begin{equation}
x(\mathbf{s};\mathbf{t}^\prime) = f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),
\end{equation}

where \(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the
parameters associated with the temporal operation and \(d_{\psi}\) is
the number of parameter of \(\psi\). A typical example of temporal
processing is aggregation and in the drought community, these indexes
are often called multi-scalar indexes. For example, Standardized
Precipitation Index (SPI) aggregates monthly precipitation using a time
scale parameter \(k\) in the form of

\[x(s_i;t_{j^\prime}) = \sum_{j = j^\prime}^{j^\prime + k - 1}x(s_i; t_j),\]

where \(j^\prime\) is the new time index after the aggregation.
Multi-scalar drought indexes are widely used to study different types of
drought events. A smaller value of the time scales parameter \(k\) is
used for short term drought events, while a large \(k\) of 18 or 24, an
equivalent to a 1.5- or 2-year aggregation, is often used for long term
assessment.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}

@@

\begin{itemize}
\item
  \(x\) is the variable in \(\mathbb{R}^p\),
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). This includes a collection of
  countries of interest and paired longitude and latitude coordinates of
  locations, and
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  Sometimes, it can be time breaks into other temporal components, such
  as year, month, quarter, and season.
\end{itemize}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1781}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
\caption{An notation overview of the input, operation, and output of
each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x(t)\) & \(f[x(.)]\) & \(x^*(t^\prime)\)
where \(t\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x(s)\) & aaaaa & \(x^\#(s^\prime)\) \\
3 & Variable transformation & \(x(s; t)\) & \(T(x(s;t))\) & \(u(s;t)\)
???? \\
4 & Scaling & \(x(s; t)\) & \([x(s;t) - \alpha]/\gamma\) & \(v(s;t)\)
where bbbbb \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\) & aaaaa &
\(\mathbf{y}(s;t)\) where \(y \in \mathbb{R}^d\) and \(d < p\) \\
6 & Distribution fit & \(x(s; t)\) & \(F_\mathbf{\theta}(x(s,t))\) where
\(F_{\mathbf{\theta}}(.)\) is the cumulated distribution function with
distribution parameter(s) \(\mathbf{\theta}\) & \(p(s,t)\) where
\(p(.)\) denotes the probability value between 0 and 1 \\
7 & Normalising & \(x(s; t)\) & \(\Phi^{-1}[x(s, t)]\) & \(z(s; t)\)
where \(z \sim N(.,.)\) \\
8 & Benchmarking & \(x(s; t)\) & & \(b(s;t)\) \\
9 & Simplification & \(x(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

\texttt{Output}: \(x^*(t^{\prime})\), where
\(t^\prime \in D_{t^{\prime}}\).

\texttt{Computation}: The temporal processing module takes a single
variable in a set of fixed locations, \(x(t)\), as the input. This step
\emph{transforms} the original series into a new series to incorporate
the temporal neighbouring effect: \(x^*(.) = f[x(.)]\). In temporal
processing, the temporal dimension \(t\) may also transition to
\(t^\prime \subseteq D_{t^\prime}\), allowing for the decomposition of
time into its constituent temporal components. For example, this can be
used to break down daily data into monthly. This gives the output of the
temporal processing module as \(x^*(t^\prime)\).

An example of temporal processing is in Standardized Precipitation Index
(SPI), where monthly precipitation is summed over a rolling window of
size \(k\), also known as the time scale. Assuming an original time
series indexed by \(t= 1, 2, \cdots, T\), the aggregated series, indexed
by \(t^\prime = 1, 2, \cdots, T-k+1\), can be written as
\(x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)\). For SPI,
the choice of the time scale \(k\), is used to control the accumulation
period for the water deficit, enabling the assessment of drought
severity across various types (meteorological, agricultural, and
hydrological).
